---
title: "Wide Receiver Projections in Fantasy Football"
output: html_document
date: '2022-06-08'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Title: Evaluating Wide Receiver Performance in Fantasy Football

Abstract: 
Fantasy football is a multi-billion dollar industry and has become a staple for casual and die-hard football fans. The use of data analysis has also exploded in the fantasy football community, mainly to make player projections so that fantasy players can gain any possible edge. Oftentimes, there are factors that cause players to exceed or fall short of their fantasy performance. The goal of this project is to look at wide receiver data to make predictions on players who could be huge values and/or busts next season based on underlying metrics. 

Context:
The stakes of Fantasy Football can be very high. Whether it is professional players playing for huge prizes or just your common player playing for bragging rights over their friends, players are heavily invested and want to win, badly. An inevitable part of Fantasy Football is that there will be players who fall very short of their expectations. On the flipside, there are players who exceed their expectations. Oftentimes, there are very obvious reasons for why one or the other happens such as injuries or a change at quarterback leading to more or less success. Other times, however, there can be more underlying issues that may have pointed to or contributed to success (or lack thereof).

The ultimate goal of Fantasy Football is to win, and have fun while doing it. There is undeniably a solid amount of “luck” involved in Fantasy Football. However, with so much data and player evaluation metrics now at our disposal, Fantasy players rely heavily on projections to help them craft their rosters. Projections are extremely hard and no one can get them right all the time. But, it is important for the Fantasy Football community to have people who are dedicated to doing their best at making accurate projections for players. 

Proposal:
I hope to investigate wide receiver data specifically. The reasoning is because they are the largest position in terms of quantity in Fantasy Football. They are also some of the most influential players, and targeting the right/wrong receivers can make or break the chances of winning your league. The main question that I am trying to answer is: What are the biggest predictors of success for wide receivers in Fantasy Football? Based on these predictors, who are some wide receivers that players should target or stay away from in drafts for next season?

I plan on using wide receiver data from past seasons to find correlations to fantasy performance in terms of points. From there, I hope to take the results from the correlation tests to identify players who performed well/poorly in these metrics and compare them with the amount of fantasy points these players scored last season. I also plan on running regression models to attempt to answer some of the same questions. My analysis will be done predominantly in R. 

I expect to find that there were a fair share of wide receivers who either outperformed their ability or vice versa. These players would then presumably be due for negative or positive regression to the mean for the upcoming season. My initial prediction is that wide receiver targets/target share percentage is going to be the biggest indicator of fantasy points scored. I also hope to be able to use what I find for my own Fantasy Football drafts in the upcoming season. 

Conclusion:
Over and under performance is often an issue in Fantasy Football. Success in one season does not guarantee success in another, and the same goes for the reverse. It is important for Fantasy Football players to have access to projections that are backed up by what the data says. I plan to use data in order to assess wide receiver performance and find receivers that Fantasy players should either target or steer clear from in the 2022 season. 

This project has the potential to be expanded in the future. I could look at all the other positions in the Fantasy Football player pool as well (Quarterbacks, Running Backs, Tight Ends). It would be interesting to look at the different metrics that predict success for other positions than wide receivers, in the future. 


EDA:

```{r}
library(tidyverse)
library(readr)
library(readxl)
library(blogdown)
```

## Importing main datasets

```{r}
fantasy_stats <- read.csv("fantasy-football-leaders_wr.csv")
advanced_wr_stats <- read.csv("advanced-wr-metrics.csv")
advanced_wr_efficiency <- read.csv("advanced-wr-efficiency-metrics.csv")
```

```{r}
head(fantasy_stats)
head(advanced_wr_stats)
head(advanced_wr_efficiency)
```

### Who averages the most touchdowns/season over the last 5 years

```{r}
fantasy_stats %>%
  filter(Season == '2021' | Season == '2020' | Season == '2019' | Season == '2018' | Season == '2017') %>%
  filter(sum(GMS) >= 20) %>%
  group_by(Name) %>%
  summarize(avg_tds = mean(REC_TD)) %>%
  arrange(desc(avg_tds)) %>%
  head(15)
```

### Who had high discrepencies between air yards and points scored last season?

```{r}
air_yds <- fantasy_stats %>%
  filter(Season == '2021') %>%
  left_join(advanced_wr_stats, by = "ID") %>%
  filter(Season.y == '2021', FPTS >= 150) %>%
  select(Name.x, FPTS, TOT_AIR_YARDS)

head(air_yds)
```


```{r}
### Looking for lots of air yards but lower fantasy finishes
air_yds[order(-air_yds$TOT_AIR_YARDS, air_yds$FPTS), ]
```

Names that stand out here are Courtland Sutton(!!), Terry McLaurin, Marvin Jones Jr., DeVonta Smith, Darnell Mooney, Chase Claypool, Emmanuel Sanders. This would presumably point to wide receivers who are receiving a lot of targets in terms of yards but are not doing a great job of turnin them into fantasy points. High air yards with lower fantasy points could be a predictor of receivers that are due for positive regression

```{r}
### What about high fantasy points with lower air yards?
air_yds[order(-air_yds$FPTS, air_yds$TOT_AIR_YARDS), ]
```

Deebo Samuel (unicorn), Hunter Renfrow, Jaylen Waddle, Chris Godwin, Amon-Ra St. Brown, Adam Thielen. This could indicate players who have a lot of shorter yard recpetions but not a lot of potential for explosive plays.

### Correlation tests

```{r}
fant_cor <- fantasy_stats %>%
  select(-c(Season, Rk, ID, Name, Team, Pos))
```

```{r}
cor(fant_cor[ , colnames(fant_cor) != "FPTS"], fant_cor$FPTS)
```

From standard fantasy statistics, Rec Yards, Rec, Targets, PPG and Rec Td's are the most correlated metrics to scoring fantasy points

## Touchdown leaders last season

```{r}
advanced_wr_stats %>%
  filter(Season == '2021') %>%
  group_by(Name) %>%
  select(Name, TD) %>%
  arrange(desc(TD)) %>%
  head(10) %>%
ggplot(aes(reorder(Name, TD), TD)) +
  geom_col(fill = 'lightblue') + 
  coord_flip() +
  theme_classic()
```

